---
title: "Gaussian Process Models"
author: "Sean Martin"
date: "September 16, 2016"
output: ioslides_presentation
runtime: shiny
---

## A quick introduction to Gaussian Process models

Some stuff

## Motivation

Why are we using these things?

## Parametric Models

Take linear models as a starting point. These model the data as

## Basis Expansion

Sometimes a linear function of the input isn't ideal. In this case, we can model non-linear functions of the data using basis expansion

- Polynomial regression
- Basis splines
- Fourier Series

##
### Polynomial Regression

##
### Basis Splines

##
### Fourier Series

## Kernel Regression

More generally, 

## Gaussian Processes

##
### 1-dimensional example:

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```


##
### 2-dimensional example:

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```


##
### Taking a break for some examples

## Other cool stuff related to GPs

##
### GP-LVMs

##
### Classification

##
### Deep GPs
